### 2015-08-11 Paper/paper.tex
Add first ideas and semi-sentences for PFLP paper

\documentclass[
12pt, % Main document font size
a4paper, % Paper type, use 'letterpaper' for US Letter paper
oneside, % One page layout (no page indentation)
]{llncs}
% \usepackage[german]{babel}
\title{Declarative Pearl: Lightweight Probabilistic Programming in Curry}
\author{Sandra Dylus}
\institute{Institut f\"ur Informatik, Christian-Albrechts-Universi\"at zu Kiel\\\email{sad@informatik.uni-kiel.de}}
\begin{document}
\maketitle

\begin{abstract}
  Lorem ipsum
\end{abstract}

\section{Introduction}

There are many probabilistic languages around these days; a progress
that started longer than a decade ago, but is currently especially
active. %
These languages have their origin either in functional, logic or
imperative programming, i.e., you can choose a language based on your
favourite paradigm. %
Besides newly developed languages, a few libraries for probabilistic
programming have seen the light of day as well. %
These languages (stand-alone or DSL) all come with a handfull of
useful tools (functions and stuff) to work with probabilistic values
and distributions as well as inference algorithms to solve
probabilistic models that use these values and distributions. %
That said, there is a large subset that all these languages have in
common, dispite their different approaches on semantic, syntax as well
as goals and range of usage, respectively. %
This subset is rather of practical nature and not of big intereset
from a PL researcher perspective: every good library or programming
language needs a handfull of primitives and predefined functions, but
they are not the heart of this piece of software. %

In case of probabilistic programming languages and libraries, we need
to look at the implementation of the actual semantic behaviour -- what
can we express with probabilistic languages and how can we interpret
such an expression? %
What is the intuition behind probability in the context of programming
and what is its meaning? %

Contributions:

\begin{itemize}
\item We present a lightweight library for probabilistic programming
  in Curry. %
  The library's core consists of ... data structures as well as
  ... functions to work on these strucutres, overall the library needs
  less than 30 lines of code. %
\item We can reduce the library code to such a minimum by reusing Curry's
  semantic. %
  The best part is: no further mechanisms are necessary to enable
  probabilistic computations in Curry. %
  That said, we show high connections between current probability
  programming languages and Curry, including lazy evaluation,
  nondeterministic values as well as corresponding built-in search
  abilities and call-time choice. %
\item We present examples that emphasise the benefits of Curry's
  representation of nondeterminsm, which enables early pruning of the
  search space, compared to Haskell's MonadPlus instance for lists,
  which is often used to model nondeterminism. %
\end{itemize}

\section{The Problem}

\section{The Idea}

\begin{itemize}

\item We treat nondeterministic values like \emph{normal} values, that
  is, we can test predicates on such values in order to filter
  unwanted results. %

\begin{verbatim}
filterDist :: (a -> Bool) -> Dist a -> Dist a
filterDist p d | p (value d) = d
\end{verbatim}

  Such a filter is often called \emph{partial identity} and a common
  functional logic design pattern \cite{funcLogPattern}, additionaly,
  it corresponds to notion of \emph{observing} a probabilistic value
  and is a standard primitive in probabilistic programming
  languages. %

\item we model probabilistic values with nondeterminstic values -- a
  probabilistic value is a pair of a value and its probability; a
  probabilistic value %

\begin{verbatim}
biasedCoin = (True, 0.3)
biasedCoin = (False, 0.7)
\end{verbatim}

\begin{verbatim}
coin = (True ? False,0.5)
\end{verbatim}

\item Each nondeterministic choice corresponds to a branch in the
  search tree. %
  When we evaluate the search tree, we start at the root of the search
  tree and go along a path in that tree: we decide for each branch
  which child to take next. %
  Additionally, we keep track of each decision we make in order to
  repeat a decision in case it occurs several times at different
  positions within the search tree. %

\begin{verbatim}
plus :: Dist Int -> Dist Int -> Dist Int
plus dI1 dI2 = (+) <$> dI1 <*> dI2

doubleCoinWithUnexpectedOutcome =
  coin `plus` coin

doubleCoin =
  let c = coin
  in c `plus` c
\end{verbatim}

\end{itemize}

\section{The Details}

\section{Related Work}

\section{Conclusion and Future Work}

\bibliography{/Users/sad/Documents/Papers/used}
\bibliographystyle{plain}
\end{document}

### 2015-08-12 Paper/paper.tex
Make further progress on draft

@@
 \section{Introduction}
-
 There are many probabilistic languages around these days; a progress
@@
 programming have seen the light of day as well. %
-These languages (stand-alone or DSL) all come with a handfull of
+These languages (stand-alone or DSL) all come with a handful of
 useful tools (functions and stuff) to work with probabilistic values
@@
 from a PL researcher perspective: every good library or programming
-language needs a handfull of primitives and predefined functions, but
+language needs a handful of primitives and predefined functions, but
 they are not the heart of this piece of software. %
@@
 
+The main goal of probabilistic programming is to open the door for a
+wide range of groups that want to evaluate probabilistic models: data
+scientists, domain experts and other developers. %
+Probabilistic programming languages want to minimise the burden for
+non-experts (in the area of machine learning and co) to design their
+own model and run algorithm on their data. %
+These languages already provide a rich set of primitives and most
+importantly inference algorithms that are built upon the results of
+experts. %
+Thus, a predefined and well-elaborated toolset simplifies the access
+to machine learning, because technical details can be ignored. %
+
 Contributions:
-
 \begin{itemize}
-\item We present a lightweight library for probabilistic programming
-  in Curry. %
+\item We present \emph{PFLP} a lightweight library for probabilistic
+  programming in the functional logic language Curry. %
   The library's core consists of ... data structures as well as
@@
 \begin{itemize}
-
-\item We treat nondeterministic values like \emph{normal} values, that
-  is, we can test predicates on such values in order to filter
-  unwanted results. %
-
-\begin{verbatim}
-filterDist :: (a -> Bool) -> Dist a -> Dist a
-filterDist p d | p (value d) = d
-\end{verbatim}
-
-  Such a filter is often called \emph{partial identity} and a common
-  functional logic design pattern \cite{funcLogPattern}, additionaly,
-  it corresponds to notion of \emph{observing} a probabilistic value
-  and is a standard primitive in probabilistic programming
-  languages. %
-
 \item we model probabilistic values with nondeterminstic values -- a
@@
 
+\item Working within a host language enables us to focus on the
+  problem itself and frees us from reimplementing standard primitives
+  (IO, data structures), essential design decisions (syntax,
+  parser). %
+  Based on which criteria do you choose the host language? %
+  As for probabilistic languages that started growing over time,
+  libraries that enable probablistic programming exist in the whole
+  variety of paradigms: declarative as well as imperative. %
+  For a start, you obviously choose the paradigm that has the clearest
+  structure and enables straight-forward programming. %
+  We want to address a wide range of developers and even
+  non-PL-experts, thus, we choose a paradigm than can be learnt
+  easily. %
+  Thus, obvioulsy, you choose a declarative programming language. %
+  By choosing Curry, we argue that it is sufficient to reuse built-in
+  features in order to cover the essence of probabilistic programs,
+  especially its semantic. %
+
+\end{itemize}
+
+\section{The Details}
+
+\begin{itemize}
+
+\item We treat nondeterministic values like \emph{normal} values, that
+  is, we can test predicates on nondeterministic values in order to
+  filter unwanted results. %
+
+\begin{verbatim}
+filterDist :: (a -> Bool) -> Dist a -> Dist a
+filterDist p d | p (value d) = d
+\end{verbatim}
+
+  Such a filter is often called \emph{partial identity} and a common
+  functional logic design pattern \cite{funcLogPattern}, additionaly,
+  it corresponds to notion of \emph{observing} a probabilistic value
+  and is a standard primitive in probabilistic programming
+  languages. %
+
 \item Each nondeterministic choice corresponds to a branch in the
@@
   positions within the search tree. %
+  However, we have to distinct between values that are choosen
+  randomly values that share the decision. %
+  This distinction is very natural in Curry: a let-binding forces to
+  share decisions on probabilistic values. %
 
@@
 
+\item We can benefit from lazy evaluation: a given search tree is only
+  evaluated if really needed and only the parts that are essential for
+  the given query, e.g., observations can rule out whole branches
+  early. %
+
 \end{itemize}
-
-\section{The Details}
 

### 2015-08-13 Paper/paper.tex
Make progress on draft

@@
 \institute{Institut f\"ur Informatik, Christian-Albrechts-Universi\"at zu Kiel\\\email{sad@informatik.uni-kiel.de}}
+
+\newcommand{\code}[1]{{\texttt{#1}}}
+
 \begin{document}
@@
 
+  Rules with overlapping patterns on the left-hand side are one way to
+  define nondeterministic functions in Curry. %
+  Alternatively, Curry provides an explicite choice operator \code{?}
+  that chooses nondeterministic between its first and second
+  argument. %
+
 \begin{verbatim}
@@
   functional logic design pattern \cite{funcLogPattern}, additionaly,
-  it corresponds to notion of \emph{observing} a probabilistic value
-  and is a standard primitive in probabilistic programming
-  languages. %
+  it corresponds to notion of observing a probabilistic value and is a
+  standard primitive in probabilistic programming languages. %
 
@@
 
+\item The choice-operator comes in handy to simplify the corresponding
+  search tree. %
+  This benefit becomes more clear with the following code examples. %
+
+\begin{verbatim}
+coin' = (True, 0.5) ? (False, 0.5)
+
+coin = (True ? False, 0.5)
+\end{verbatim}
+
+  There is no semantic difference between the first and second
+  definition, however, the corresponding search trees indeed differ. %
+
+  Choice (True, 0.5) (False, 0.5)\\%
+  (Choice True False, 0.5)\\%
+
+  The definition of \code{coin'} consists of an expression with a
+  nondeterminstic choice on top-level position, whereas the definition
+  of \cote{coin} generalises the common parts of the two possible
+  values and uses the nondeterminism in a deeper position. %
+
 \item We can benefit from lazy evaluation: a given search tree is only

### 2015-08-14 Paper/paper.tex
Add introductory example for bayesian network in Curry

--- 2015-08-12 Paper/paper.tex
@@
 \section{The Problem}
+
+Experts often state that you do not need to be a rocket scientist in
+order to make sense of research results; however, in the area of
+machine learning, beginners often face difficulties to use the tools
+and current knowledge for their own project. %
+That is, beginners need to inverse many hours to digest technical
+details before they can actually make use of top-of-the-art models and
+algorithms. %
+
+In the field of programming languages, the community often tries to
+generalise and simplify commen concepts and repetative structures. %
+Fortunately, the rising collaboration between ML and PL communities
+has produced a number of programming languages and libraries for
+probabilistic programming that aim at minimising the obstacle to use
+practicle results of ML, thus, empower more users to apply ML
+techniques to their problems. %
+
+For instance, a typical problem in the field of \emph{Bayesian
+  Network}s is to transfer a given model (e.g., a conditional
+probability table) into code in order to evaluation the model for a
+given data set. %
+Such an evaluation can range from stating queries to documenting
+observations to apply learning algorithms. %
+
+\begin{figure}
+\caption{Bayesian network with conditional probability table for \emph{alarm}-scenario}
+\end{figure}
+
+We can model this bayesian network with the following code. %
+
+\begin{verbatim}
+earthquake = bernoulli 0.2
+burglary   = bernoulli 0.7
+
+alarm True  True  = bernoulli 0.9
+alarm True  False = bernoulli 0.8
+alarm False True  = bernoulli 0.1
+alarm False False = bernoulli 0.0
+\end{verbatim}
+
+The definition for the probabilistic functions \code{earthquake},
+\code{burglary} and \code{alarm} follow straight-forward from the
+scheme of the given model. %
+The function \code{bernoulli} yields a probabilistic value that is
+\code{True} with probability given by the first argument and
+\code{False} with probability $1 - prob$. %
+
+\begin{verbatim}
+bernoulli prob = mkDist True prob
+bernoulli prob = mkDist False (1.0 - prob)
+\end{verbatim}
 

### 2015-08-14 Paper/paper.tex
Start with section on the idea of PFLP

@@
 \code{True} with probability given by the first argument and
-\code{False} with probability $1 - prob$. %
+\code{False} with the converse probability. %
 
@@
 
+After defining the model, we want to ask queries about certian
+propoerties and probabilities, respectively. %
+The essence of probabilistic programming is to harmonise the problem
+and its analysis: we want to define both, the problem and the queries,
+in same language. %
+
 \section{The Idea}
 
-\begin{itemize}
-\item we model probabilistic values with nondeterminstic values -- a
-  probabilistic value is a pair of a value and its probability; a
-  probabilistic value %
+In the style of \cite{pfp}, the library's heart is a data type that
+represents distributions: such a distribution consists of pair of
+value and its probability. %
 
 \begin{verbatim}
-biasedCoin = (True, 0.3)
-biasedCoin = (False, 0.7)
+data Probability = Prop Float
+data Dist a = Dist (a,Probability)
+\end{verbatim}
+
+However, the attentive reader may notice a slight difference: the PFLP
+approach does not need to represent the collection of possible values
+and their corresponding probabilities as list. %
+In functional languages like Curry, the representation of several
+results is commonly represented with nondeterminism instead of
+directly using a container like lists. %
+That is, we model probabilistic values as nondeterministic pairs of
+possible value tagged with their probability. %
+
+For example, we can define a biased coin with a $30\%-70\%$ chance to
+land on head and tail, respectively, as follows. %
+
+\begin{verbatim}
+data Coin = Head | Tail
+biasedCoin = (Head, 0.3)
+biasedCoin = (Tail, 0.7)
 \end{verbatim}
@@
 Alternatively, Curry provides an explicite choice operator \code{?}
-  that chooses nondeterministic between its first and second
+that chooses nondeterministicly between its first and second
 argument. %
+Hence, we can a define a fair coin with the following rule. %
 
 \begin{verbatim}
-coin = (True ? False,0.5)
+fairCoin = (Head ? Tail,0.5)
 \end{verbatim}
 
+\begin{itemize}
 \item Working within a host language enables us to focus on the
@@
   nondeterminstic choice on top-level position, whereas the definition
-  of \cote{coin} generalises the common parts of the two possible
+  of \code{coin} generalises the common parts of the two possible
   values and uses the nondeterminism in a deeper position. %

### 2015-08-18 Paper/paper.tex
Start with quering a bayesian network example

@@
 In the style of \cite{pfp}, the library's heart is a data type that
-represents distributions: such a distribution consists of pair of
-value and its probability. %
+represents distributions: such a distribution consists of a value and
+its probability. %
 
@@
 data Probability = Prop Float
-data Dist a = Dist (a,Probability)
+data Dist a = Dist a Probability
+
+mkDist :: a -> Probability -> Dist a
+mkDist = Dist
 \end{verbatim}
@@
 However, the attentive reader may notice a slight difference: the PFLP
-approach does not need to represent the collection of possible values
-and their corresponding probabilities as list. %
-In functional languages like Curry, the representation of several
-results is commonly represented with nondeterminism instead of
+approach does not need to represent a distribution as a list of its
+possible values. %
+In functional logic languages like Curry, the representation of
+several results is commonly represented with nondeterminism instead of
 directly using a container like lists. %
@@
 
-For example, we can define a biased coin with a $30\%-70\%$ chance to
-land on head and tail, respectively, as follows. %
+For example, we can define a biased coin with a $30\%$ vs $70\%$
+chance to land on head and tail, respectively, as follows. %
 
 \begin{verbatim}
-data Coin = Head | Tail
-biasedCoin = (Head, 0.3)
-biasedCoin = (Tail, 0.7)
+type Coin = Bool
+
+head, tail :: Coin
+head = True
+tail = False
+
+biasedCoin = (head, 0.3)
+biasedCoin = (tail, 0.7)
 \end{verbatim}
@@
 \begin{verbatim}
-fairCoin = (Head ? Tail,0.5)
+fairCoin = (head ? tail,0.5)
 \end{verbatim}
 
-\begin{itemize}
-\item Working within a host language enables us to focus on the
-  problem itself and frees us from reimplementing standard primitives
-  (IO, data structures), essential design decisions (syntax,
-  parser). %
-  Based on which criteria do you choose the host language? %
-  As for probabilistic languages that started growing over time,
-  libraries that enable probablistic programming exist in the whole
-  variety of paradigms: declarative as well as imperative. %
-  For a start, you obviously choose the paradigm that has the clearest
-  structure and enables straight-forward programming. %
-  We want to address a wide range of developers and even
-  non-PL-experts, thus, we choose a paradigm than can be learnt
-  easily. %
-  Thus, obvioulsy, you choose a declarative programming language. %
-  By choosing Curry, we argue that it is sufficient to reuse built-in
-  features in order to cover the essence of probabilistic programs,
-  especially its semantic. %
+Let us now take a look at the bayesian network example again. %
+We have already seen the definition of the model in the
+introduction. %
+However, the more interesting part is to make queries on the given
+model that cannot be answered with the probability table itself. %
+For example, we can ask the model for the probability that a burglary
+occured given the fact that the alarm was set of. %
 
-\end{itemize}
+\begin{verbatim}
+-- P(B=T | A=T)
+alarmByBurglary =
+  burglary >>>= \b ->
+  earthquake >>>= \e ->
+  alarm b e >>>= \a ->
+  guard a >>>= \_ ->
+  pure b
+\end{verbatim}
 
@@
 
+\item Working within a host language enables us to focus on the
+  problem itself and frees us from reimplementing standard primitives
+  (IO, data structures), essential design decisions (syntax,
+  parser). %
+  Based on which criteria do you choose the host language? %
+  As for probabilistic languages that started growing over time,
+  libraries that enable probablistic programming exist in the whole
+  variety of paradigms: declarative as well as imperative. %
+  For a start, you obviously choose the paradigm that has the clearest
+  structure and enables straight-forward programming. %
+  We want to address a wide range of developers and even
+  non-PL-experts, thus, we choose a paradigm than can be learnt
+  easily. %
+  Thus, obvioulsy, you choose a declarative programming language. %
+  By choosing Curry, we argue that it is sufficient to reuse built-in
+  features in order to cover the essence of probabilistic programs,
+  especially its semantic. %
+
 \end{itemize}

### 2015-08-27 Paper/paper.tex
Change type of mkDist
Clean up wet grass example and add debugger tools as contributions

@@
 
-The main goal of probabilistic programming is to open the door for a
-wide range of groups that want to evaluate probabilistic models: data
+The main goal of probabilistic programming is to open the door to the
+evaluation of probabilistic models for a wide range of users: data
 scientists, domain experts and other developers. %
@@
 non-experts (in the area of machine learning and co) to design their
-own model and run algorithm on their data. %
+own model and run inference algorithms on their data. %
 These languages already provide a rich set of primitives and most
@@
   which is often used to model nondeterminism. %
+\item It is fair to assume that a wide range of tools existing for
+  Curry can be applied to probabilistic languaeges as well. %
+  For instance, \cite{Brassel-BIO} has already presented work on a
+  debugger for Curry that works well with nondeterminism and free
+  variables. %
 \end{itemize}
@@
 \begin{figure}
-\caption{Bayesian network with conditional probability table for \emph{alarm}-scenario}
+\label{fig:WetGrass}
+\caption{Bayesian network with conditional probability table for
+  \emph{Wet Grass}-scenario}
 \end{figure}
 
-We can model this bayesian network with the following code. %
+We can model the bayesian network given in Figure \ref{fig:WetGrass} with the following code. %
 
 \begin{verbatim}
-earthquake = bernoulli 0.2
-burglary   = bernoulli 0.7
+rain :: Dist Bool
+rain = bernoulli 0.2
 
-alarm True  True  = bernoulli 0.9
-alarm True  False = bernoulli 0.8
-alarm False True  = bernoulli 0.1
-alarm False False = bernoulli 0.0
+sprinkler :: Bool -> Dist Bool
+sprinkler False = bernoulli 0.4
+sprinkler True  = bernoulli 0.01
+
+grassWet :: Bool -> Bool -> Dist Bool
+grassWet False False = bernoulli 0.0
+grassWet False True  = bernoulli 0.8
+grassWet True  False = bernoulli 0.9
+grassWet True  True  = bernoulli 0.99
 \end{verbatim}
 
-The definition for the probabilistic functions \code{earthquake},
-\code{burglary} and \code{alarm} follow straight-forward from the
+The definition for the probabilistic functions \code{rain},
+\code{sprinkler} and \code{grassWet} follow straight-forward from the
 scheme of the given model. %
@@
 \begin{verbatim}
-bernoulli prob = mkDist True prob
-bernoulli prob = mkDist False (1.0 - prob)
+bernoulli prob = dist True prob
+bernoulli prob = dist False (1.0 - prob)
 \end{verbatim}
@@
 
-mkDist :: a -> Probability -> Dist a
-mkDist = Dist
+dist :: a -> Float -> Dist a
+dist x prob = Dist x (Prob prob)
 \end{verbatim}
@@
 
-biasedCoin = (head, 0.3)
-biasedCoin = (tail, 0.7)
+biasedCoin = dist head 0.3
+biasedCoin = disy tail 0.7
 \end{verbatim}
@@
 \begin{verbatim}
-fairCoin = (head ? tail,0.5)
+fairCoin = dist (head ? tail) 0.5
 \end{verbatim}
@@
 model that cannot be answered with the probability table itself. %
-For example, we can ask the model for the probability that a burglary
-occured given the fact that the alarm was set of. %
+For example, we can ask the model for the probability that it was
+raining when observing wet grass. %
 
@@
 -- P(B=T | A=T)
-alarmByBurglary =
-  burglary >>>= \b ->
-  earthquake >>>= \e ->
-  alarm b e >>>= \a ->
-  guard a >>>= \_ ->
-  pure b
+grassIsWet =
+  rain >>>= \r ->
+  sprinkler r >>>= \s ->
+  grassWet s r = >>>= \g ->
+  guard g >>>= \_ ->
+  pure r
+\end{verbatim}
+
+\begin{verbatim}
+guard :: Bool -> Dist ()
+guard True = pure ()
 \end{verbatim}
@@
 \begin{verbatim}
-coin' = (True, 0.5) ? (False, 0.5)
+coin' = mkDist True 0.5 ? mkDist False 0.5
 
-coin = (True ? False, 0.5)
+coin = mkDist (True ? False) 0.5
 \end{verbatim}

### 2015-08-31 Paper/paper.tex
Add ideas for each chapter

@@
 \section{Introduction}
-There are many probabilistic languages around these days; a progress
-that started longer than a decade ago, but is currently especially
-active. %
-These languages have their origin either in functional, logic or
-imperative programming, i.e., you can choose a language based on your
-favourite paradigm. %
-Besides newly developed languages, a few libraries for probabilistic
-programming have seen the light of day as well. %
-These languages (stand-alone or DSL) all come with a handful of
-useful tools (functions and stuff) to work with probabilistic values
-and distributions as well as inference algorithms to solve
-probabilistic models that use these values and distributions. %
-That said, there is a large subset that all these languages have in
-common, dispite their different approaches on semantic, syntax as well
-as goals and range of usage, respectively. %
-This subset is rather of practical nature and not of big intereset
-from a PL researcher perspective: every good library or programming
-language needs a handful of primitives and predefined functions, but
-they are not the heart of this piece of software. %
+\begin{itemize}
+\item more and more PPL
+\item all paradigms are represented
+\item (E)DSLs as well
+\item all languages have common ground (similar primitives and
+  predefined helper functions)
+\item implementation is the heart of these ``software products'' --
+  what can be expressed?
+\item main goal of PPLs: tool for a wide range of researchers and not
+  only ML experts; minimise burden for beginners; no former background
+  on inference algorithms needed, define only model and ask queries;
+\end{itemize}
 
-In case of probabilistic programming languages and libraries, we need
-to look at the implementation of the actual semantic behaviour -- what
-can we express with probabilistic languages and how can we interpret
-such an expression? %
-What is the intuition behind probability in the context of programming
-and what is its meaning? %
+% There are many probabilistic languages around these days; a progress
+% that started longer than a decade ago, but is currently especially
+% active. %
+% These languages have their origin either in functional, logic or
+% imperative programming, i.e., you can choose a language based on your
+% favourite paradigm. %
+% Besides newly developed languages, a few libraries for probabilistic
+% programming have seen the light of day as well. %
+% These languages (stand-alone or DSL) all come with a handful of
+% useful tools (functions and stuff) to work with probabilistic values
+% and distributions as well as inference algorithms to solve
+% probabilistic models that use these values and distributions. %
+% That said, there is a large subset that all these languages have in
+% common, dispite their different approaches on semantic, syntax as well
+% as goals and range of usage, respectively. %
+% This subset is rather of practical nature and not of big intereset
+% from a PL researcher perspective: every good library or programming
+% language needs a handful of primitives and predefined functions, but
+% they are not the heart of this piece of software. %
 
-The main goal of probabilistic programming is to open the door to the
-evaluation of probabilistic models for a wide range of users: data
-scientists, domain experts and other developers. %
-Probabilistic programming languages want to minimise the burden for
-non-experts (in the area of machine learning and co) to design their
-own model and run inference algorithms on their data. %
-These languages already provide a rich set of primitives and most
-importantly inference algorithms that are built upon the results of
-experts. %
-Thus, a predefined and well-elaborated toolset simplifies the access
-to machine learning, because technical details can be ignored. %
+% In case of probabilistic programming languages and libraries, we need
+% to look at the implementation of the actual semantic behaviour -- what
+% can we express with probabilistic languages and how can we interpret
+% such an expression? %
+% What is the intuition behind probability in the context of programming
+% and what is its meaning? %
+
+% The main goal of probabilistic programming is to open the door to the
+% evaluation of probabilistic models for a wide range of users: data
+% scientists, domain experts and other developers. %
+% Probabilistic programming languages want to minimise the burden for
+% non-experts (in the area of machine learning and co) to design their
+% own model and run inference algorithms on their data. %
+% These languages already provide a rich set of primitives and most
+% importantly inference algorithms that are built upon the results of
+% experts. %
+% Thus, a predefined and well-elaborated toolset simplifies the access
+% to machine learning, because technical details can be ignored. %
 
@@
 
-Experts often state that you do not need to be a rocket scientist in
-order to make sense of research results; however, in the area of
-machine learning, beginners often face difficulties to use the tools
-and current knowledge for their own project. %
-That is, beginners need to inverse many hours to digest technical
-details before they can actually make use of top-of-the-art models and
-algorithms. %
+\begin{itemize}
+\item abstraction from actual ML problems: define only model and
+  questions
+\item give simple example for Bayesian Networks (Wet Grass, Alarm,
+  etc) and explain functions as superficial as possible
+\item point out resemblence between CPT and actual code for model
+\end{itemize}
+% Experts often state that you do not need to be a rocket scientist in
+% order to make sense of research results; however, in the area of
+% machine learning, beginners often face difficulties to use the tools
+% and current knowledge for their own project. %
+% That is, beginners need to invest many hours to digest technical
+% details before they can actually make use of top-of-the-art models and
+% algorithms. %
 
-In the field of programming languages, the community often tries to
-generalise and simplify commen concepts and repetative structures. %
-Fortunately, the rising collaboration between ML and PL communities
-has produced a number of programming languages and libraries for
-probabilistic programming that aim at minimising the obstacle to use
-practicle results of ML, thus, empower more users to apply ML
-techniques to their problems. %
+% In the field of programming languages, the community often tries to
+% generalise and simplify commen concepts and repetative structures. %
+% Fortunately, the rising collaboration between ML and PL communities
+% has produced a number of programming languages and libraries for
+% probabilistic programming that aim at minimising the obstacle to use
+% practicle results of ML, thus, empower more users to apply ML
+% techniques to their problems. %
 
@@
 After defining the model, we want to ask queries about certian
-propoerties and probabilities, respectively. %
-The essence of probabilistic programming is to harmonise the problem
-and its analysis: we want to define both, the problem and the queries,
-in same language. %
+properties and probabilities, respectively. %
+The essence of probabilistic programming is to harmonise a problem and
+its analysis: we want to define both, the problem and the queries, in
+same language. %
 
 \section{The Idea}
+
+\begin{itemize}
+\item PPL vs DSL\\
+  Working within a host language enables us to focus on the
+  problem itself and frees us from reimplementing standard primitives
+  (IO, data structures), essential design decisions (syntax,
+  parser). %
+  % Based on which criteria do you choose the host language? %
+  % As for probabilistic languages that started growing over time,
+  % libraries that enable probablistic programming exist in the whole
+  % variety of paradigms: declarative as well as imperative. %
+  % For a start, you obviously choose the paradigm that has the clearest
+  % structure and enables straight-forward programming. %
+  % We want to address a wide range of developers and even
+  % non-PL-experts, thus, we choose a paradigm than can be learnt
+  % easily. %
+  % Thus, obvioulsy, you choose a declarative programming language. %
+  % By choosing Curry, we argue that it is sufficient to reuse built-in
+  % features in order to cover the essence of probabilistic programs,
+  % especially its semantic. %
+\item describe underlying implementation: core data types and
+  functions
+\item query simple question for example from introduction
+\item connect to Martin Erwig's PFP approach
+\item Nondeterminism and \code{SearchTree}
+\end{itemize}
 
@@
 
-\item Working within a host language enables us to focus on the
-  problem itself and frees us from reimplementing standard primitives
-  (IO, data structures), essential design decisions (syntax,
-  parser). %
-  Based on which criteria do you choose the host language? %
-  As for probabilistic languages that started growing over time,
-  libraries that enable probablistic programming exist in the whole
-  variety of paradigms: declarative as well as imperative. %
-  For a start, you obviously choose the paradigm that has the clearest
-  structure and enables straight-forward programming. %
-  We want to address a wide range of developers and even
-  non-PL-experts, thus, we choose a paradigm than can be learnt
-  easily. %
-  Thus, obvioulsy, you choose a declarative programming language. %
-  By choosing Curry, we argue that it is sufficient to reuse built-in
-  features in order to cover the essence of probabilistic programs,
-  especially its semantic. %
-
 \end{itemize}
@@
 
+\begin{itemize}
+\item Martin Erwig (FPL)
+\item Oleg Kiselyov (HANSEI)
+\item ProbLog (on top of Prolog), AILog (debugger), Church (functional language)
+\end{itemize}
+
 \section{Conclusion and Future Work}
+\begin{itemize}
+\item SetFunctions (memory allocation; usage as ``first-class
+  citizen'', i.e., check semantics)
+\item ``outside'' nondeterminism
+\item sampling technique for nondeterministic functions in Curry
+\item reevaluate Currys semantic in constrast to Sato's distributuion semantic
+\item reactivate debugger for KiCS2/PAKCS
+\end{itemize}
 

### 2015-09-09 Paper/paper.tex
Add comments to conclusion (disadvantage in contrast to HANSEI and in general corresponding to sampling VS exact inference)

@@
 % That said, there is a large subset that all these languages have in
-% common, dispite their different approaches on semantic, syntax as well
+% common, despite their different approaches on semantic, syntax as well
 % as goals and range of usage, respectively. %
-% This subset is rather of practical nature and not of big intereset
+% This subset is rather of practical nature and not of big interest
 % from a PL researcher perspective: every good library or programming
@@
   The library's core consists of ... data structures as well as
-  ... functions to work on these strucutres, overall the library needs
+  ... functions to work on these structures, overall the library needs
   less than 30 lines of code. %
@@
 \item We present examples that emphasise the benefits of Curry's
-  representation of nondeterminsm, which enables early pruning of the
+  representation of nondeterminism, which enables early pruning of the
   search space, compared to Haskell's MonadPlus instance for lists,
@@
 \item It is fair to assume that a wide range of tools existing for
-  Curry can be applied to probabilistic languaeges as well. %
+  Curry can be applied to probabilistic languages as well. %
   For instance, \cite{Brassel-BIO} has already presented work on a
@@
   etc) and explain functions as superficial as possible
-\item point out resemblence between CPT and actual code for model
+\item point out resemblance between CPT and actual code for model
 \end{itemize}
@@
 % In the field of programming languages, the community often tries to
-% generalise and simplify commen concepts and repetative structures. %
+% generalise and simplify common concepts and repetitive structures. %
 % Fortunately, the rising collaboration between ML and PL communities
@@
 % probabilistic programming that aim at minimising the obstacle to use
-% practicle results of ML, thus, empower more users to apply ML
+% practical results of ML, thus, empower more users to apply ML
 % techniques to their problems. %
@@
 
-After defining the model, we want to ask queries about certian
+After defining the model, we want to ask queries about certain
 properties and probabilities, respectively. %
@@
   % As for probabilistic languages that started growing over time,
-  % libraries that enable probablistic programming exist in the whole
+  % libraries that enable probabilistic programming exist in the whole
   % variety of paradigms: declarative as well as imperative. %
@@
   % easily. %
-  % Thus, obvioulsy, you choose a declarative programming language. %
+  % Thus, obviously, you choose a declarative programming language. %
   % By choosing Curry, we argue that it is sufficient to reuse built-in
@@
 Alternatively, Curry provides an explicite choice operator \code{?}
-that chooses nondeterministicly between its first and second
+that chooses nondeterministic between its first and second
 argument. %
@@
   Such a filter is often called \emph{partial identity} and a common
-  functional logic design pattern \cite{funcLogPattern}, additionaly,
+  functional logic design pattern \cite{funcLogPattern}, additionally,
   it corresponds to notion of observing a probabilistic value and is a
@@
   positions within the search tree. %
-  However, we have to distinct between values that are choosen
-  randomly values that share the decision. %
+  However, we have to distinct between values that are chosen randomly
+  values that share the decision. %
   This distinction is very natural in Curry: a let-binding forces to
@@
   The definition of \code{coin'} consists of an expression with a
-  nondeterminstic choice on top-level position, whereas the definition
-  of \code{coin} generalises the common parts of the two possible
-  values and uses the nondeterminism in a deeper position. %
+  nondeterministic choice on top-level position, whereas the
+  definition of \code{coin} generalises the common parts of the two
+  possible values and uses the nondeterminism in a deeper position. %
 
@@
 \item Oleg Kiselyov (HANSEI)
-\item ProbLog (on top of Prolog), AILog (debugger), Church (functional language)
+\item ProbLog (on top of Prolog), AILog (debugger), Church (functional
+  language)
 \end{itemize}
@@
 \section{Conclusion and Future Work}
+\begin{itemize}
+\item search tree can only manage weights at leaves, hence, sampling
+  is not a feasible alternative to exact inference (that is really
+  bad!)
+\item Furthermore, a big disadvantage of using Curry in contrast to
+  HANSEI is the representation of the search tree. %
+  Because of the built-in nondeterminism, we did not have to
+  reimplement such a mechanism, but we also have to stick to the
+  implementation at is. %
+  HANSEI's representation of the search tree has the advantage of
+  weighted branches, whereas Curry can only simulate distributions as
+  weighted leaves within the search tree. %
+
+\begin{verbatim}
+HANSEI
+              1.0
+        /            \
+      0.5            0.5
+    /     \        /     \
+0.25     0.25    0.3     0.2
+\end{verbatim}
+
+\begin{verbatim}
+Curry
+               .
+         /          \
+       .             .
+    /     \       /      \
+0.25     0.25   0.3      0.2
+\end{verbatim}
+
+  That is, for exact inference, both approaches have to explore the
+  whole search tree, but sampling becomes a lot more easier in HANSEI,
+  because at each branch we can prune the search tree by choosing one
+  branch and getting rid of all the others. %
+  In Curry, we cannot do that, because we do not have any information
+  about the probabilities, so in the worst-case we have to explore the
+  whole search tree to sample a value. %
+
+\item 
+\begin{quote}
+  Another point of difference will come out when we talk about memo. %
+  We need some sort of state, and that state cannot be global state. %
+  When the search is encapsulated, that state has to be be
+  encapsulated too, somehow. %
+  Hansei also supports so-called nested inference, or nested
+  encapsulated search, in Curry terms. %
+  That is, when you encapsulate search, some choices are encapsulated
+  (represented as nodes in a tree) and some choices are performed. %
+  I'm not sure Curry has support for such things. \\%
+
+--- mail correspondence with Oleg Kiselyov
+\end{quote}
+\end{itemize}
+
 \begin{itemize}
@@
 \item sampling technique for nondeterministic functions in Curry
-\item reevaluate Currys semantic in constrast to Sato's distributuion semantic
+\item reevaluate Currys semantic in constrast to Sato's distribution
+  semantic
 \item reactivate debugger for KiCS2/PAKCS

### 2015-09-15 Paper/paper.tex
Fix typo semantic/semantics

@@
 \item We can reduce the library code to such a minimum by reusing Curry's
-  semantic. %
+  semantics. %
   The best part is: no further mechanisms are necessary to enable
@@
   % features in order to cover the essence of probabilistic programs,
-  % especially its semantic. %
+  % especially its semantics. %
 \item describe underlying implementation: core data types and
@@
 \item sampling technique for nondeterministic functions in Curry
-\item reevaluate Currys semantic in constrast to Sato's distribution
-  semantic
+\item reevaluate Currys semantics in constrast to Sato's distribution
+  semantics
 \item reactivate debugger for KiCS2/PAKCS
